%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~%
%                                    APPENDIX TEMPLATE                                    %
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~%

% Use these after the \appendix command at the end of report.tex
% Appendices work similar to a chapter and are used to provide supplementary information that typically do not contribute to the final page count of a report.

\chapter{Long Appendix Title Here} % Main appendix title

\label{AppendixA} % for referencing this appendix elsewhere, use \ref{AppendixA}

% This is for the header on each page if in use - input a shortened title due to constrictions.
% \lhead{Appendix A. \emph{Short App. Title Here}}

\section{Gauge Freedom in GR (NR)}

One of the most important characteristic of GR is its so called diffeomorphism invariance which essentially means
that the "physics" inside theory is unchanged under smooth coordinate transformations. In other words, different coordinate systems (or gauges) can describe the same underlying spacetime geometry.


Mathematically, this invariance implies that if $g_{\mu\nu}(x)$ is a solution to Einstein's equations, then so is the transformed metric under a diffeomorphism $x^\mu \rightarrow x'^\mu(x)$, with the metric transforming as
\begin{equation}
g'_{\mu\nu}(x') = \frac{\partial x^\alpha}{\partial x'^\mu} \frac{\partial x^\beta}{\partial x'^\nu} g_{\alpha\beta}(x).
\end{equation}

Often in theoretical studies we utilise various gauges to uncover the internal dynamics of the spacetime variables. The easiest example would in the case of gravitational waves where we often assume a linearised gravity ($g_{\mu\nu}= \eta_{\mu \nu} + h_{\mu \nu}$ where $O(h) << 1$) where the Minkowskian flat space metric is considered as the background metric of the spacetime.  Here to retrieve the notion of gravitational waves we use something known as Lorenz gauge or De Donder gauge ($\partial^{\mu} \bar{h}_{\mu\nu}$) where $\bar{h}$ is the trace reversed metric perturbation. 
Using this gauge we will be able to see that, the metric perturbation has formalised into the form of a wave equation.

$$
\Box \bar{h}^{\mu\nu}= -16 \pi T^{\mu \nu}
$$
This is finest and easiest example of utilising gauges to understand the internal physics. But in the case of the NR, it is utilised differently. As
we know in NR, we try to see Einstein's equations as a set of PDE's to solve numerically. But to solve a problem numerically it is essential that the problem we are solving is \textit{well-posed} and devoid of any coordinate artifacts that coordinate system may evolve into(coordinate singularities , constraint violation and so on). So essentially, we utilise the gauge freedom to counter these problems. Based on how we tackle them has given rise to various formulations of NR like BSSN, Z4C etc. In our case we use the Generalised Harmonic Formulation of NR.

\section{Generalised Harmonic Gauge}

The main advantage of the Generalised Harmonic gauge is that it is, by construction, a well-posed formulation of Einstein's equations. This makes it particularly suitable for numerical evolution, as it provides a strongly hyperbolic system with well-defined characteristic structure. Due to this property, practical considerations aside, in theory they are ideal for any kind of numerical evolutions performed with GR in mind.  The well-posedness of this formulation comes from the trace reversed form of Einstein's equation i.e



\begin{equation}
R_{\mu\nu}
= -\frac{1}{2} g^{\alpha\beta} \partial_\alpha \partial_\beta g_{\mu\nu}
+ \partial_{(\mu} \Gamma_{\nu)}
+ \Gamma^\alpha \Gamma_{\mu\nu\alpha}
- \Gamma^\alpha_{\mu\beta} \Gamma^\beta_{\nu\alpha}
\label{GHG}
\end{equation}

where $
\Gamma^\mu = g^{\alpha\beta} \Gamma^\mu_{\alpha\beta}$ called the contracted christoffel symbols. The well-posedness or ill-posedness of a PDE is often determined by its principal terms (Terms containing the highest derivatives). In our problem, the principal terms are the first and the second part of the Equation \ref{GHG} since they contain the second derivatives of the metric. But since the wave equation is a well posed problem, the problematic part of the equation is the second term (derivative of contracted christoffels). But for that we can utilise the gauge freedom to ensure that they don't contain second derivatives of the metric i.e
$\Gamma^a=\nabla_b \nabla^b x^a=-H^a(x,g)$ where $H^a$ is called the Gauge Source function and it is a function of atmost the metric and not its derivatives.

This gives us the following equations of motion for the lapse($\alpha$) and shift($\beta$)

\begin{equation}
\partial_t \alpha = -\alpha^2 (K + H_n) + \mathcal{L}_\beta \alpha
\qquad \partial_t \beta^i = \alpha^2 (\Gamma^i + \perp H^i) - \alpha D^i \alpha + \beta^j \partial_j \beta^i
\label{lapse=shift}
\end{equation}
where $H^n= n^a H_a$

The $H^a$'s are freely specifiable functions but as we mentioned before, we choose it in a way to have proper evolutions without any coordinate pathologies. That choice will be mentioned in the coming sections.

\section{Harmonic Horizon Marching Gauge}
\label{HHMG}
Since GR has an intrinsic gauge freedom built into it, it will be reflected when we do simulations using NR as well. Hence we will have to choose a gauge but in the case of simulations they are chosen by the stability it confers to the simulations.

In \texttt{BAMPS}, the usual gauge that we use is HDWG (Harmonic Gauge). Here the evolution equations for the lapse becomes,

\[
\mathcal{L}_n \log \alpha = - (H_n +K)
\]

where $H_n$ is the free choosable function which only contains the metric and not its derivatives.

Inorder the push the lapse to 1 we add an extra term into the harmonic function in such a way that it adds a term $\eta_{hm}(\alpha-1)$ into the evolution equation for lapse

\[
\partial_t \alpha -\beta^i \partial_i \alpha= -\alpha^2(H_n + K)
\]

The extra term that in $H_n$ is $\eta_{hm} \alpha^{-2} (\alpha-1)$

\[
\partial_t \alpha -\beta^i \partial_i \alpha= -(\eta_{hm} (\alpha-1) + \alpha^2 K)
\]

So in $H_a$ an extra term of $-\eta_{hm} \alpha^{-2} (\alpha - 1) n_a$ is added

So New $H_0=.... + \eta_{hm} \alpha^{-1} (\alpha-1)$

\[
\partial_t H_0 = +\eta_{hm} \alpha^{-2} \partial_t \alpha
\]

\[
\partial_i H_0 = +\eta_{hm} \alpha^{-2} \partial_i \alpha
\]