%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~%
%	                                     ABSTRACT   	                                  %
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~%
\begin{flushleft} % This section is not essential for the abstract
    \setlength{\parskip}{0pt}
        %\bigskip
    {\centering{{\Large{\bf{ABSTRACT}}}} \par}
    \medskip
    \vspace{6pt}
    \hrule \vspace{0.4cm}
		  Student Name:	 \textbf{\ThesisAuthor}	\hfill Reg. No.: \textbf{\RegNumber} \\
		Degree:  \textbf{\DegreeDisplay} \hfill Dept.:  \textbf{School of \Department} \\
		Thesis Title: \textbf{\ThesisTitle}\\
		Thesis Supervisor:  \textbf{\ProjectSupervisor}\\
		Date of thesis submission: \textbf{{ \today}\\[0.4cm] }
	\hrule \vspace{1.5cm} 
\end{flushleft}
\noindent
Critical phenomena in gravitational collapse constitute one of the earliest major results in numerical relativity, revealing universal behavior at the threshold of black hole formation. In this work, we investigate the emergence of such phenomena in axisymmetric vacuum spacetimes with twist (which encodes the rotational degree of freedom of the spacetime). By constructing and evolving one-parameter families of twisting Brill wave initial data, we study the threshold between dispersion and black hole formation. Near the critical threshold, we analyze scaling behavior using curvature invariants and examine indications of universality and self-similarity. Our results show evidence of power-law scaling and suggest the presence of echo-like features, although higher resolution is required for conclusive identification. We also identify key numerical challenges, including gauge effects, coordinate pathologies, and limitations in horizon-finding techniques, which become significant near criticality. This work provides insight into critical phenomena in vacuum spacetimes beyond spherical symmetry and highlights the role of twist in gravitational collapse.
\vspace{4cm}

{\large \textbf{Keywords: }}\par{\Large \ThesisKeywords}

\cleardoublepage